---
title: "BST209 Final Project Overview"
author: "Diana Juvinao-Quintero"
date: "7/30/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Project Overview

The National Health and Nutrition Examination Survey(NHANES) assesses the health and nutritional statusof adults and children in the United States.The goal is to use these covariates to predict the occurrence of diabetes, cancer, heart attack, stroke,OR depression.
   
## Research Question
   
Identify a predictive model for the risk of diabetes based on a combination of dietary, demographic, lab and examination factors  from the NHANES US national survey using Machine Learning methods.
   
## Data Source
   
The National Health and Nutrition Examination Survey (NHANES) assesses the health and nutritional status of adults and children in the United States. The `covariate_df` data frame in the `covariate_df.rData` file contains the demographic, examination, diet, and lab data from [NHANES 2013-2014](https://www.kaggle.com/datasets/cdc/national-health-and-nutrition-examination-survey). You can peruse the dictionaries for the variables on the CDC website: [demographics](https://wwwn.cdc.gov/Nchs/Nhanes/Search/variablelist.aspx?Component=Demographics&CycleBeginYear=2013), 
[examinations](https://wwwn.cdc.gov/Nchs/Nhanes/Search/variablelist.aspx?Component=Examination&CycleBeginYear=2013), 
[diet](https://wwwn.cdc.gov/Nchs/Nhanes/Search/variablelist.aspx?Component=Dietary&CycleBeginYear=2013), and
[lab](https://wwwn.cdc.gov/Nchs/Nhanes/Search/variablelist.aspx?Component=Laboratory&CycleBeginYear=2013). 

Your goal is to use these covariates to predict the occurrence of diabetes, cancer, heart attack, stroke, OR depression. The company wants you to recommend a predictive model and identify key variables that appear to be associated with the outcome disease. 
   
## Methods
* Load data 
* select outcome and covariates and perform data cleaning and processing
* Visualize the relationship between covariates and outcome (4-10 outputs)
* Build machine 3 learning model
   
## Expected Outcomes
   
[Describe what you expect to find]


### Loading data and libraries
As part of data processing, choose ONE of the five diseases and merge the corresponding variable from the `questionnaire.csv` dataset with `covariate_df`. This disease is the outcome that your model should predict. The relevant variable names in `questionnaire.csv` are: DIQ010 (diabetes), MCQ220 (cancer), MCQ160E (heart attack), MCQ160F (stroke), and DPQ020 (depression). (You can view the entire questionnaire dictionary [here](https://wwwn.cdc.gov/Nchs/Nhanes/Search/variablelist.aspx?Component=Questionnaire&CycleBeginYear=2013).) For each disease variable, you can take `1` to be affected, `2` to be unaffected, and all other numeric values to be unknown/ignored. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
library(caret)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(tableone)
library(knitr)
```

```{r}
rm(list=ls())

load("covariate_df.rData")
outcomes<-read.csv("questionnaire.csv")

```

### Selecting outcome and covariates
```{r}
outcomes[1:5, 1:10]
head(colnames(outcomes))

covariate_df[1:5, 1:10]
head(colnames(covariate_df))

# merge two datasets
data<-left_join(covariate_df, outcomes %>% select(SEQN, DIQ010), by="SEQN") # selecting diabetes as outcome
dim(data) # 9813

# select covariates of interest

## country, educ (2), age(y),gender, time in US, total house income
demographics<-c("DMDHRBR4","DMDEDUC2","DMDHREDU", "DMDHRAGE","DMDHRGND","DMDYRSUS", "INDHHIN2") 

## diastolic 3rd read, systolic 3rd read, DXA abdominal diameter, BMI (kg/m2), waist circum (cm), overall oral health
examination<-c("BPXDI3","BPXSY3", "BMDAVSAD", "BMXBMI", "BMXWAIST", "OHDEXSTS")

dietary<-c()
lab<-c()

```


### Data cleaning
1. Perform any necessary data pre-processing and cleaning, and document your steps. Depending on the prompt you selected, this may involve transforming variables, creating new variables, and merging data frames. In particular, you may need to make some decisions on how to handle missing data, such as removing rows or columns with a significant amount of missing observations, creating an "unknown" category, or replacing/imputing the missing values. You do not need to develop a rigorous process or cite references, but please briefly justify your choices. 


```{r}

```


### Data visualization
2. Make and interpret 4-10 visualizations to help you understand the relationships between the variables in your dataset. We highly encourage you to explore the data on your own, but when preparing your response to this question, please be parsimonious in your plots and select visualizations that help you tell a story about the data. If you need to make additional plots to support your responses to the other questions (e.g. to motivate data cleaning or modeling choices), that's fine. 

```{r}

```


### Model selection

3. Build any 3 machine learning models that have been presented in class that use your choice of covariates to predict the given outcome variable. Explain why you chose those covariates and interpret the model performances. If you are including categorical variables as covariates in a `glmnet` model (lasso or ridge regression), please remember that you will need to convert your covariate data frame into a *model matrix*, e.g. by calling the `model.matrix` function:

```{r}
model_matrix = model.matrix(~.-1, my_covariate_df)
```

